| Early stopping in clinical PET studies: \n| how to reduce expense and exposure

### 2020-10-15 ArticleDraft/Draft.Rmd
Updated suppl and some typos in main text

$H_0: \delta = 0$
$H_1: \delta \sim Cauchy(0,0.707)$

@@

\begin{figure}
        \centering
        \includegraphics[width=140mm]{/Users/pontus.sigray/Github/Early_stopping_in_PET/Results/Figures/Figure4.png}
        \caption{\textit{Panels A and B show the proportion of studies that showed support for H\textsubscript{1} (aka “power curves”) for BF sequential testing (blue) and fixed N test (black), at different population effects (starting at D = $0$, i.e. no effect). For the fixed N approach, only one test is performed at N = $30$ subjects/group. For the sequential testing, $12$ subjects/group are first collected, then BF is checked after each added comparison pair until $30$ subjects/group is reached, using a stopping threshold of $4$. Panels C and D show the average number of subjects needed to reach a stopping decision at different population effects. The fixed N test is the black line (N = $30$ subjects/group); BF sequential testing is the blue line with shaded area denoting $\pm$ $1$SD.}}
        \label{fig_powerPhase1}
\end{figure}

@@

The area around the point where the red and black line cross[^fn7], is a weak spot for the BF sequential testing procedure. At this range of true effects in the population, the risk of obtaining an inconclusive result is at its highest (i.e., the blue line is at its nadir). In the cases in which a decision is reached, the study will stop with equal probability for H~1~ and H~0~, and the risk for false negative evidence is around 25%.

Assuming that the PET researcher would be interested in stopping a study early when H~0~ is supported, a range of population effects with high risk of false negative evidence will always exist. It is therefore sensible to beforehand decide on a minimal population effect of interest, and choose the settings to ensure that any larger effect does not have too high a risk of generating evidence for H~0~. For example, with the settings presented in Figure \ref{fig_powerPhase2} (Nstart = 12, Nmax = 30, threshold = 4) it could be said that all effects between D = 0 and 0.45 ought to be of little clinical interest, in order for sequential BF testing to be applicable in this case since the risk of false negative evidence for these effects is > 10%[^fn8].

@@

###Methods
We applied sequential BF testing to assess the support in data in favor of patients having higher BP~F~ in the raphe nuclei than controls (H~1~), as compared to no difference (H~0~). We also examined the evidence in data in favor of H~0~ over H~1~, to stop the study early if the null hypothesis was supported. The stopping threshold was set to BF>5 for H~1~ (BF < $\nicefrac{1}{5}$ for H~0~), in order to keep the rate of false positive evidence below 5%[^fn9] (see Supplementary Figure S1). We used a one-sided Cauchy distribution, centered around 0 with an *r* = 0.707 to describe the predicted mean difference under H~1~. H~0~ was specified as the point zero value.

@@

MDD.tidy <- read.csv('/Users/pontus.sigray/Github/Early_stopping_in_PET/RawData/MDD_WAY_data/MDD_way_tidy_NHST_results.csv')

@@

MDD.d <- abs(round(MDD.tidy$cohensD,2))
MDD.perc <- abs(round(MDD.tidy$percDiff,2))*100

@@

###Results
When applying a NHST one-sided two-sample t-test to the full dataset, we observed a large group difference in raphe nuclei BP~F~ (t = `r MDD.tval`, df = `r MDD.df`, p = `r MDD.pval`), with patients showing higher values compared to healthy controls (D = `r MDD.d` or a `r MDD.perc`% increase).

@@

#Discussion
In this tutorial we show that it is possible to stop recruitment of subjects in PET studies as soon as enough data have been collected to make a conclusion, saving both expense and exposure to radioactivity. We do this by employing sequential BF testing, which assesses the support in data in favor of two competing hypotheses while the study is still ongoing.

@@

It should by now be clear to the reader that a set of critical decisions needs to be made before applying sequential BF testing in a clinical PET study. For this reason, we recommend all researchers to pre-register their analysis before data collection starts [^fn10] [@VantVeer2016; @Poldrack2017; @Knudsen2020]. A pre-registration can be thought of as a safety net for the researcher. It helps guide the analysis and interpretation of data so that error rates are kept under control. It can also be shown to reviewers or readers to increase the credibility of the methods and findings. Deviations from a pre-registration is of course possible, and often warranted, but should be reported transparently in the article.

@@

[^fn7]:Egon: Don’t cross the streams. \newline 
Peter: Why? \newline 
Egon: It would be bad. \newline 
Peter: I’m fuzzy on the whole good/bad thing. What do you mean “bad”? \newline 
Egon: Try to imagine all life as you know it stopping instantaneously and every molecule in your body exploding at the speed of light. \newline 
Raymond: Total protonic reversal. \newline 
Peter: That’s bad. Okay. Alright, important safety tip, thanks Egon. \newline 
*- Ghost Busters, 1984* \newline \phantom{x}

[^fn8]: We believe that it is reasonable to consider a false negative to be somewhat less severe than a false positive,  and for this reason we choose to present a 10% threshold here. However, just as with the conventional 5% threshold for false positive findings [@Lakens2018], this is not a sacrosanct value. We encourage the reader to think for themselves what proportion of false evidence they are ready to accept when planning their own clinical PET studies.  \newline \phantom{x}

[^fn9]:A H~0~ stopping threshold of BF <$\nicefrac{1}{5}$ means that if there is a true population difference between patients and controls in the range of D = 0 to 0.35, the risk of a false negative finding will be >10%. Hence, if the researchers consider such effects to be of high clinical interest to detect, they might consider using other settings instead. See the discussion for possible solutions. \newline \phantom{x}

[^fn10]: See e.g., https://aspredicted.org/ for easy and efficient pre-registration. \newline \phantom{x}

>>>>

$H_0: \delta = 0$

$H_1: \delta \sim Cauchy(0,0.707)$

@@

\begin{figure}
        \centering
        \includegraphics[width=140mm]{/Users/pontus.sigray/Github/Early_stopping_in_PET/Results/Figures/Figure4.png}
        \caption{\textit{Panels A and B show the proportion of studies that showed support for H\textsubscript{1} (aka “power curves”) for BF sequential testing (blue) and fixed N test (black), at different population effects (starting at D = $0$, i.e. no effect). For the fixed N approach, only one test is performed at N = $30$ subjects/group. For the sequential testing, $12$ subjects/group are first collected, then BF is checked after each added comparison pair until $30$ subjects/group is reached, stopping the study when BF > $4$. Panels C and D show the average number of subjects needed to reach a stopping decision at different population effects. The fixed N test is the black line (N = $30$ subjects/group); BF sequential testing is the blue line with shaded area denoting $\pm$ $1$SD.}}
        \label{fig_powerPhase1}
\end{figure}

@@

The area around the point where the red and black line cross, is a weak spot for the BF sequential testing procedure. At this range of true effects in the population, the risk of obtaining an inconclusive result is at its highest (i.e., the blue line is at its nadir). In the cases in which a decision is reached, the study will stop with equal probability for H~1~ and H~0~, and the risk for false negative evidence is around 25%.

Assuming that the PET researcher would be interested in stopping a study early when H~0~ is supported, a range of population effects with high risk of false negative evidence will always exist. It is therefore sensible to beforehand decide on a minimal population effect of interest, and choose the settings to ensure that any larger effect does not have too high a risk of generating evidence for H~0~. For example, with the settings presented in Figure \ref{fig_powerPhase2} (Nstart = 12, Nmax = 30, threshold = 4) it could be said that all effects between D = 0 and 0.45 ought to be of little clinical interest, in order for sequential BF testing to be applicable in this case since the risk of false negative evidence for these effects is > 10%[^fn7].

@@

###Methods
We applied sequential BF testing to assess the support in data in favor of patients having higher BP~F~ in the raphe nuclei than controls (H~1~), as compared to no difference (H~0~). We also examined the evidence in data in favor of H~0~ over H~1~, to stop the study early if the null hypothesis was supported. The stopping threshold was set to BF>5 for H~1~ (BF < $\nicefrac{1}{5}$ for H~0~), in order to keep the rate of false positive evidence below 5%[^fn8] (see Supplementary Figure S1). We used a one-sided Cauchy distribution, centered around 0 with an *r* = 0.707 to describe the predicted mean difference under H~1~. H~0~ was specified as the point zero value.

@@

MDD.tidy <- read.csv('/Users/pontus.sigray/Github/Early_stopping_in_PET/RawData/MDD_WAY_data/MDD_way_tidy_fixedN_results.csv')

@@

MDD.d <- abs(round(MDD.tidy$cohensD,2))
MDD.perc <- abs(round(MDD.tidy$percDiff,2))*100
MDD.BF <- abs(round(MDD.tidy$BF))

@@

###Results
When applying a one-sided two-sample t-test to the full dataset, we observed a large group difference in raphe nuclei BP~F~ (t = `r MDD.tval`, df = `r MDD.df`, p = `r MDD.pval`, BF = `r MDD.BF` ), with patients showing higher values compared to healthy controls (D = `r MDD.d` or a `r MDD.perc`% increase).

@@

#Discussion
In this tutorial we show that it is possible to stop recruitment of subjects in PET studies as soon as enough data have been collected to make a conclusion, reducing both expense and exposure to radioactivity. We do this by employing sequential BF testing, which assesses the support in data in favor of two competing hypotheses while the study is still ongoing.

@@

It should by now be clear to the reader that a set of critical decisions needs to be made before applying sequential BF testing in a clinical PET study. For this reason, we recommend all researchers to pre-register their analysis before data collection starts [^fn9] [@VantVeer2016; @Poldrack2017; @Knudsen2020]. A pre-registration can be thought of as a safety net for the researcher. It helps guide the analysis and interpretation of data so that error rates are kept under control. It can also be shown to reviewers or readers to increase the credibility of the methods and findings. Deviations from a pre-registration is of course possible, and often warranted, but should be reported transparently in the article.

@@

[^fn7]: We believe that it is reasonable to consider a false negative to be somewhat less severe than a false positive,  and for this reason we choose to present a 10% threshold here. However, just as with the conventional 5% threshold for false positive findings [@Lakens2018], this is not a sacrosanct value. We encourage the reader to think for themselves what proportion of false evidence they are ready to accept when planning their own clinical PET studies.  \newline \phantom{x}

[^fn8]:A H~0~ stopping threshold of BF <$\nicefrac{1}{5}$ means that if there is a true population difference between patients and controls in the range of D = 0 to 0.35, the risk of a false negative finding will be >10%. Hence, if the researchers consider such effects to be of high clinical interest to detect, they might consider using other settings instead. See the discussion for possible solutions. \newline \phantom{x}

[^fn9]: See e.g., https://aspredicted.org/ for easy and efficient pre-registration. \newline \phantom{x}
